A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-27 main.tex
Various small fixes
Closes: #43
Closes: #903
Closes: #32
Evaluation is now a separate file

We baseline using the same Presburger solver (\Princess{}), input file parser,
and automaton implementation as \Catra. We do this in order to better analyse
the impact of the calculus rules themselves. Using the formula of
\eqref{eq:generate-parikh}, we produce quantified Presburger formulae for each
successive term and add them to \Princess. We compute the product incrementally
term by term, checking satisfiability at each step. We use a priority queue to
select automata for each step, and order it by the number of transitions as a
heuristic for the size of the automata. We use this heuristic to avoid computing
large (and therefore slow) products until we have to, banking our hopes on
computing an empty intermittent product early. The pseudocode for our
implementation can be seen in \cref{alg:baseline}.

@@

\section{Evaluation}\label{sec:experiments}

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances of Parikh
automata intersection problems generated by the \OstrichPlus{} string constraint
solver when solving the PyEx benchmarks~\cite{pyex}. After generating an
initial~\InitialNrBenchmarks{},we remove~\NrTrivial{} instances solved in under
five seconds by baseline. We also attempted to benchmark instances generated by \Ostrich{} solving the kaluza-len (\numprint{38227} instances) and pyex-len (\numprint{791} instances) benchmark suites, but had to discard them since they were all trivial. For instance, on a laptop, \Calculus{} could solve every instance from kaluza-len in under five seconds, mean about 0.1s. The benchmarks are run on commit~\texttt{\commit}.

The benchmarks are executed in parallel using GNU Parallel~\cite{parallel},
since they are mostly single-threaded and initial results showed negligible
interference on performance, on \BenchmarkRig{}. We compiled the code using
Scala~\ScalaVersion{}, and executed the experiments on~\JvmVersion{} with a
maximum heap of~\MaxHeapSize{}. We used \Nuxmv{} version~\NuxmvVersion{} invoked
as a subprocess for each instance. Instances were executed in batches of
\BatchSize{}, each given a fresh JVM. We believe this represents a realistic use
case where \Calculus{} is used to support e.g a string solver. Experiments were
executed in random order for all backends. Each instance got a time budget
of~\RuntimeTimeout.

All runtimes are measured in wall-clock time as observed by the JVM when
executing the instance, and exclude time spent parsing (usually far below
\numprint{0.1}s).

\subsection{Execution Time and Ability to Solve Instances}\label{sec:runtime}

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances left after excluding the trivial instances, among them the \numprint{38227} and \numprint{791} instances generated from kaluza-len and pyex-len respectively,
the respective back-ends could solve and with which status. A full summary of
their outcomes is also available in \cref{tab:solve-status}. Note that we lack
ground truth for all instances. We see that \Calculus{} generally outperforms
\Nuxmv{} on determining unsatisfiability, while being worse at satisfiable
instances. Both \Nuxmv{} and \Calculus{} strictly outperform baseline on every
kind of instance, but most of all on satisfiable instances, and \Calculus{}
outperforms \Nuxmv{}. \Nuxmv{} and \Calculus{} disagree on 38 instances, where
\Nuxmv{} returns \Sat{} and \Calculus{} returns \Unsat{}. We believe this to be
caused by a non-deterministically triggered bug in the clause learning of
\Calculus{}, but have not had time to further investigate it. We have however
validated all \numprint{3908} satisfying assignments from \Calculus{} with
\Nuxmv{}. \Cref{tab:solve-status} also shows a small number of crashing bugs in
our implementation.

Baseline performs worse on satisfiable instances because it executes a heuristic
meant to detect unsatisfiability early at a heavy penalty to satisfiable
instances. The heuristic is enabled since it is unrealistic to assume that we
know in advance whether an instance is satisfiable or not, but we do know that
unsatisfiable instances are more common.

\begin{table}
  \centering
  \input{graphs/solved_pivot_table.tex}
  \caption{The result of running the respective back-ends with a timeout of
  \RuntimeTimeout. Instances solved by no backend within the timeout are omitted
  from the table. }\label{tab:solve-status}
\end{table}

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-by-solver.pdf}
    \caption{The division of statuses per backend.}
    \label{fig:solve-division}
    \Description[A bar chart showing three bars, one per backend, illustrating how many instances they could solve]{The bars are divided by satisfiable and unsatisfiable instances. Baseline could seemingly only solve unsatisfiable instances, lazy could solve a few satisfiable and mostly unsatisfiable, and nuxmv could solve about twice as many unsatisfiable as satisfiable instances, and slightly less in total than lazy.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-time-boxplot.pdf}
    \caption{The distribution of runtimes for solved instances per backend. Note that the number of instances solved differs between backends.}
    \label{fig:runtime-boxplot}
    \Description[A box plot showing the distribution of runtime over the three backends]{The middle box plot shows a tiny box centered around 0 seconds for the lazy backend, the rightmost box shows a bigger box between five seconds and 30 seconds for nuxmv, and the leftmost box shows a smaller box between 20 and 30 seconds for baseline. The whiskers of nuxmv span between 0 and the timeout, 60 seconds, while they are much tighter for lazy. On the other hand, lazy has  a lot of outliers.}
  \end{subfigure}
  \vfill
  \begin{subfigure}[b]{0.75\textwidth}
    \includegraphics[width=\textwidth]{graphs/\commit-cactus.pdf}
    \caption{The number of instances solved as the time budget increases, simulated from one \RuntimeTimeout-timeout run.}
    \label{fig:cactus}
    \Description[A cactus plot comparing the performance of nuxmv, lazy and baseline]{The line for baseline is strictly lower than the two others, and does not head upwards from zero until after 10 seconds of timeout, while the number of instances for nuxmv increases sharply until ca 10 seconds and then linearly after that. Lazy solves much more instances than nuxmv.}
  \end{subfigure}
\end{figure}

\subsubsection{Scalability}\label{sec:scaling}

A cactus plot showing the number of instances solved within a given timeout for
each backend can be seen in \cref{fig:cactus}. Here, we see clearly that
\Calculus{} outperforms \Nuxmv{} in general, but that \Nuxmv{} might have better
long-term scaling properties.

\iffalse
\subsection{Finding a Presburger Formula}\label{sec:evaluation:finding-image}

For baseline and \Calculus{}, \Catra{} offers the ability to find the equivalent Presburger formula representing a given instance. For baseline, we use the built-in quantifier elimination facilities of the underlying \Princess{} theorem prover, while for \Catra{} we use the specially tailored approach described in \cref{sec:finding-the-image}. For this experiment, we use only the~\NrKnownSat{} instances known to be satisfiable from the previous experiment detailed in \cref{sec:scaling,sec:runtime}.

To make sure baseline puts up as much competition as possible, we disable
checking intermittent satisfiability and configure \Catra{} to run in the
maximally eager mode where the product is first computed before any
satisifiability check is performed. We run the experiments with a timeout
of~\ImageTimeout{}. The results of the experiment is summarised in
\cref{fig:cactus:image} and \cref{tab:image-results}. \Fudge{We see here that
something happens}.

\begin{figure}[ht]
  \caption{The number of instances \Catra{} was able to find the Presburger form of the image for within a given number of seconds per backend.}
  \label{fig:cactus:image}
\end{figure}
\fi

\subsection{Threats to Validity}

The most obvious threats to validity would be poor benchmarking or poor
implementation, e.g. if the method described in \cref{sec:calculus,sec:multiple}
deviate from what is actually benchmarked in section \cref{sec:experiments}, or
if the methods used for benchmarking would be unsound.  To increase the
probability that our results are representative both in the sense of
representing  on expected inputs and in the statistical sense for a given run of
\Catra{} despite the use of randomness in our implementation we execute many
experiments. Moreover, to address the issue of correctness we have validated all
reported solutions made by \Calculus{} with \Nuxmv{} to ensure that \Calculus{}
is indeed sound.

Another threat to validity would be if the competition (baseline and \Nuxmv)
would be disadvantaged in our comparison. For \Nuxmv{} we use the default
configuration which we believe should be performant (or it should not be the
default). Additionally, tweaking our invocation of \Nuxmv{} is explicitly made
easy for artefact reviewers. For baseline, our best argument is that the
solution we use is close to the one found in the \Ostrich{} string solver and
that it therefore should be realistic.

The greatest threat to validity is our choice of implementation platform and
automata library. There are some signs that product computation is inefficient,
notably the good performance of low-threshold automata materialisation that
prioritises computing smaller products. This means that \Calculus{} makes less
heavy use of product computation than baseline does, instead relying more on
\Princess{}. This situation would unfairly advantage \Calculus{} if our
automata library was the bottleneck due to our automata library being
unnaturally poor. We believe this is unlikely since similar performance issues
have been reported for string solvers. Additionally, profiling suggests that
both \Princess{}-based backends spend most of their time in \Princess{},
suggesting that the automaton implementation is not the bottleneck.

>>>>

We baseline using the same Presburger solver (\Princess{}), input file parser,
and automaton implementation as \Catra. We do this in order to better analyse
the impact of the calculus rules themselves. Using the formula of
\eqref{eq:generate-parikh}, we produce quantified Presburger formulae for each
successive term and add them to \Princess. We compute the product incrementally
term by term, checking satisfiability at each step. We use a priority queue to
select automata for each step, and order it by the number of transitions as a
heuristic for the size of the automata. We use this heuristic to avoid computing
large (and therefore slow) products until we have to, banking our hopes on
computing an empty intermittent product early.

@@

\section{Evaluation}\label{sec:experiments}
\input{evaluation.tex}
